---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals throughout the day. The data consists of two months of data from an anonymous individual collected during the months of October and November 2012, and includes the number of steps taken in 5 minute intervals each day.

## Data

The data for this assignment can be downloaded from the course web site:

- Dataset: [Activity monitoring data [52K]](https://d396qusza40orc.cloudfront.net/repdata/data/activity.zip)

The variables included in this dataset are:

- steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)
- date: The date on which the measurement was taken in YYYY-MM-DD format
- interval: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

## Loading and preprocessing the data

```{r echo=F, results='hide', message=F, warning=F}
## install dplyr if not already done so
if (("dplyr" %in% (installed.packages())) == F) {
        install.packages("dplyr")
}
## install stringr if not already done so
if (("stringr" %in% (installed.packages())) == F) {
        install.packages("stringr")
}
```

We're going to use libraries dplyr and stringr for our data processing:
```{r message=F, warning=F}
library(dplyr)
library(stringr)
```

The following code downloads, unzips, and loads the data file into a data frame:

```{r}
## the name of the dataset file
data_file <- "activity.csv"

## download dataset file if not already done so
if (!file.exists(data_file)) {
        ## check OS
        is_windows <- .Platform$OS.type == "windows"
        ## download the dataset
        download.file(url = "https://d396qusza40orc.cloudfront.net/repdata/data/activity.zip", 
                      destfile = "activity.zip",
                      method = ifelse(is_windows, "auto", "curl"))
        ## unzip the dataset archive
        unzip(zipfile = "activity.zip", files = data_file)
}

## read the entire dateset
rawdata <- read.csv(file = data_file, colClasses = c("integer", "Date", "character"))

## print out the column names of the data
names(rawdata)

## check the structure of the data
str(rawdata)

## check data for NA values
na_steps <- sum(is.na(rawdata$steps))
na_dates <- sum(is.na(rawdata$date))
na_intervals <- sum(is.na(rawdata$interval))
```

We have `r ifelse(na_steps == 0, 'no', na_steps)` NA values for steps, `r ifelse(na_dates == 0, 'no', na_dates)` NA values for dates, and `r ifelse(na_intervals == 0, 'no', na_intervals)` NA values for intervals.

When looking at the values of the column 'interval' we can observe that they are numerical with a certain pattern:
```
0, 5, 10, 15, 20, 25, 30, 35, 40, 45, 50, 55, 100...
```
The pattern seems as follows: every twelve values perform a step of 100: 
```
0, 5, ..., 100, 105, ..., 200, 205, etc.  
```
This lets us conclude that the values represent the time of the day:
```
0 for 0:00h, 5 for 0:05h, ..., 100 for 1:00h, 105 for 1:05h, etc.
```
Following this pattern we can transform the values to datetime format:

```{r}
## add a column containing datatime values using '1970-01-01' as a dummy date,
## and the transformed time of the column interval
interval_to_datetime <- function(interv) {
        digits = strsplit(str_pad(interv, 4, pad = "0"),"")[[1]]
        t_string = paste0(paste(digits[1:2], collapse = ""), ":", paste(digits[3:4], collapse = ""), ":00")
        dt_string = paste("1970-01-01", t_string)
        dt = as.POSIXct(strptime(dt_string, "%Y-%m-%d %H:%M:%S", tz = Sys.timezone()))
        dt
}
rawdata$interv_time <- sapply(rawdata$interval, interval_to_datetime)
```

## What is mean total number of steps taken per day?

1. Calculate the total number of steps taken per day

```{r}
total_steps_by_day <- 
        rawdata %>%
        na.omit %>% 
        group_by(date) %>% 
        summarise(steps = sum(steps))
```

2. Make a histogram of the total number of steps taken each day

```{r}
hist(total_steps_by_day$steps, xlab = "Total Steps Per Day", ylab = "Number of Days", 
     main = "Frequency of Total Steps per Day")
```

3. Calculate and report the mean and median of the total number of steps taken per day

```{r echo=FALSE}
mean_steps <- mean(total_steps_by_day$steps)
median_steps <- median(total_steps_by_day$steps)
```

- The mean of the total number of steps taken per days is `r format(mean_steps, nsmall = 0)`.
- The median of the total number of steps taken per days is `r format(median_steps, nsmall = 0)`.

## What is the average daily activity pattern?

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
avg_steps_per_interval <- 
        rawdata %>% 
        na.omit %>%
        group_by(interv_time) %>% 
        summarise(avg_steps = mean(steps))

plot(avg_steps_per_interval$interv_time, avg_steps_per_interval$avg_steps, type = "l",
     xlab = "Interval", ylab = "Average Steps per Interval", main = "Average Daily Activity Pattern")
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r echo=FALSE}
interval_max_steps <- avg_steps_per_interval$interv_time[which.max(avg_steps_per_interval$avg_steps)]
datetime_max_steps <- as.POSIXct(interval_max_steps, origin = "1970-01-01", tz = Sys.timezone())
datetime_max_steps <- strftime(datetime_max_steps, format="%H:%M:%S", tz = Sys.timezone())
```

The 5-minute interval at `r datetime_max_steps` contains the maximum number of steps on average accross all the days in the dataset.

## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
